Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation

### 2013-02-14 manuscript/manuscript.tex
Update calls to use the all indels sim

\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets}

@@

\begin{abstract}
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computations to cloud or commodity servers. MapReduce is becoming a standard framework for distributing computation across such compute clusters. We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance equal to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. Our implementation and source files are available at \url{http://sonmezsysbio.org/software/cloudbreak}.
\end{abstract}

@@

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets. See \textcite{Alkan:2011p547} for a review. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment, and then clustering the discordant mappings to find SVs with support from multiple pairs. Read-depth (RD) approaches, in contrast, consider the distribution of concordantly mapped reads to infer the presence of SVs from the depth of coverage of locations along the genome \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

The first RP methods used only unambiguously discordantly mapped read pairs in their analyses. However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}. Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.

@@

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate the development of an RP-based SV algorithm, Cloudbreak, for discovering genomic deletions, which computes local features based on the insert sizes of spanning mappings. Cloudbreak computes features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm. We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.

@@

The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system for storing large data sets across a cluster. In addition to a MapReduce programming framework, Hadoop also provides a distributed file system (HDFS), which facilitates storing large data sets redundantly across a cluster. MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.

The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based approaches in MapReduce. This is especially true because of the graph data structures and global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In genomic sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at varying locations along the genome and then integrating those features into SV predictions in a post-processing step. Here we show that this technique can be easily translated to the MapReduce programming model.

Our formulation divides processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes features from the set of reads that are relevant to each of a set of small intervals covering the genome. In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals. The MapReduce algorithm is summarized below with a detailed explanation following.

\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textrm{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textrm{ValidAlignmentPairs}(a_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2> \in ap$}
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure
    \State $\textrm{StructuralVariationCalls } sv \gets \textsc{PostProcess }(\Phi)$ 
  \end{algorithmic}

The alignment job uses sensitive alignment tools and settings to discover as many mapping locations for each read pair as possible. In the mapping phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair. In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found. Our framework could use any aligner that is capable of reporting multiple alignments for reads. The results we report below were generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop. Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R_1 = \left\{r_{1,1},r_{1,2},\ldots,r_{1,M}\right\}$ and $R_2 = \left\{r_{2,1},r_{2,2},\ldots,r_{2,M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,K}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We leave three functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
 \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
\end{flalign*}

The first function, \textsc{Loci}, maps an alignment to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of SpanningPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection. Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.

We describe one possible set of functions below; however, we believe that there are many possible mappings that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage, heterogeneity of the sample, etc.

@@

We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak. 
\begin{description}
\item[\sc{Loci}] Because we are detecting deletions, we map SpanningPairInfo from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs which map more than 25kb apart, or in the incorrect orientation, to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus. Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We rank the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. To estimate the probability that the mapping is correct, we approximate the probability $p^{k}_c$ of the alignment of each end being correct by:

@@

We fit the means and variances of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both to $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$. We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$. We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.

@@

We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP methods Breakdancer \autocite{Chen:2009p3} and GASVPro, \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY. \autocite{Rausch:2012he} In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).

We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been discovered by another prediction from the same method. Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length. It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.

@@

As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous deletions from the genome of J. Craig Venter \autocite{Levy:2007fb}. Since there are relatively few heterozygous deletions annotated in that set, we randomly assigned each deletion to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence. We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}. We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence. Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous deletions.

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{CHR2SIM_ROC_NEW.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
\label{chr2roc}
\end{figure}

>>>>

\title{Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation}

@@

\begin{abstract}
The detection of genomic structural variations remains one of the the most difficult challenges in analyzing high-throughput sequencing data. Recent approaches have demonstrated that considering multiple mappings of all reads, rather than only uniquely mapped discordant fragments, can improve the performance of read-pair based detection methods. However, the computational requirements for storing and processing data sets with multiple mappings can be formidable. Meanwhile, the growing size and number of sequencing data sets have led to intense interest in distributing computation to cloud or commodity servers. MapReduce is becoming a standard framework for distributing processing across such compute clusters. We describe a novel conceptual framework for structural variation detection algorithms in MapReduce based on computing local features along the genome. In this framework, we develop and evaluate a distributed deletion-finding algorithm based on fitting a Gaussian mixture model to the distribution of mapped insert sizes spanning each location in the genome. On simulated and real data sets, our approach achieves performance similar to or better than a variety of popular structural variation detection algorithms, including read-pair, split-read, and hybrid approaches, and performs well across a wide range of deletion sizes. In particular, our algorithm excels at discovering deletions of size 40bp-100bp in repetitive regions of the genome. In addition, our algorithm can accurately genotype heterozygous and homozygous deletions from diploid samples. Our implementation and source code are available at \url{http://sonmezsysbio.org/software/cloudbreak}.

\medskip
\noindent\textbf{Keywords:} genomic structural variation; distributed computing; copy number variation; high-throughput sequencing; genotyping.
\end{abstract}

@@

Popular SV detection algorithms use three main signals present in high-throughput sequencing data sets. See \textcite{Alkan:2011p547} for a review. Read-pair (RP) based methods use the distance between and orientation of the mappings of the sequenced ends of DNA fragments to identify the signatures of SVs \autocite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into vthose that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation of the fragment. These approaches then cluster the discordant mappings to find SVs with support from multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome to infer the presence of SVs \autocite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints within individual reads by mapping portions of the read to different genomic locations \autocite{Wang:2011p1607,Ye:2009p2}.

The first RP methods used only unambiguously discordantly mapped read pairs in their analyses. However, it has been recently demonstrated that including multiple mappings of discordant read pairs in the analysis can improve sensitivity in repetitive regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and thus be able to detect smaller events \autocite{Marschall:2012ek}. Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, in part because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.

@@

In this paper, we describe a framework for solving SV detection problems in Hadoop and MapReduce based on the computation of local features along the genome from paired end mappings. We then demonstrate the development in this framework of an algorithm, Cloudbreak, for discovering genomic deletions. Cloudbreak computes local features that build on the ideas of \autocite{Lee:2009da}, who modeled the distribution of insert sizes at each genomic location as a Gaussian Mixture Model (GMM). Use of the Hadoop/MapReduce framework enables scalable implementations of this class of algorithm. We then characterize our algorithm's performance on simulated and real data sets and compare it with those of recently published methods.

@@

The MapReduce programming model \autocite{Dean:2008p277} divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The MapReduce framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values. Hadoop is an open source project of the Apache Foundation which provides an implementation of the MapReduce programming framework as well as a distributed file system (HDFS) for distributing the redundant storage of large data sets across a cluster. MapReduce and Hadoop allow efficient processing of large data sets by scheduling tasks for execution as close as possible to the nodes which hold the required data, minimizing network traffic and I/O contention.

The need to separate logic into mappers and reducers makes it difficult to implement traditional RP-based SV detection approaches in MapReduce. This is especially true because of the global clustering of paired end mappings at the heart of many RP approaches. MapReduce algorithms, by contrast, excel at conducting many independent calculations in parallel. In sequencing applications, for example, MapReduce based SNP-callers Crossbow \autocite{Langmead:2009p1225} and GATK \autocite{McKenna:2010p1051} perform independent calculations for each location or partition in the genome. The RP-based SV callers MoDIL \autocite{Lee:2009da} and forestSV \autocite{Michaelson:2012fj} attempt to solve the SV detection problem by computing scores or features at locations along the genome and then producing SV predictions from those features in a post-processing step. Here we show that this technique can be easily translated to the MapReduce programming model.

Our formulation divides processing into two distinct MapReduce jobs. The first job finds possible alignment locations for pairs of reads in FASTQ format. The second job computes features from the set of alignment pairs that are relevant to each of a set of small intervals covering the genome. In a final third stage executed outside of MapReduce, the features for each location are exported from the Hadoop file system and post-processed to produce a set of SV calls which span multiple genomic intervals. This MapReduce framework is summarized below with a detailed explanation following.

\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure
    \State $\textrm{StructuralVariationCalls } svs \gets \textsc{PostProcess}(\phi_{1,2,\ldots})$ 
  \end{algorithmic}

The alignment job uses sensitive mapping tools and settings to discover as many mapping locations for each read pair as possible. In the map phase, mappers align reads in single-end mode to the reference genome in parallel, outputting each possible mapping location as a value under a key identifying the read pair. In the reduce phase, the framework combines pairs of mapping locations for the two ends of a read pair to produce all possible combinations of mapping locations found. Our framework could use any aligner that is capable of reporting multiple alignments for reads. The results we report below were generated with alignments from RazerS 3 \autocite{Weese:2012by}, which in our tests displayed the best combination of sensitivity, performance, and memory usage suitable for parallelization in Hadoop. Our implementation also contains wrappers to execute the aligners Novoalign\footnote{\url{http://wwww.novocraft.com}}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We then leave three functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
 \textsc{PostProcess} :& \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\} 
\end{flalign*}

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection. Finally, the \textsc{PostProcess} function maps the complete set of features for each locus across the genome into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$.

We describe one possible set of functions below; however, we believe that there are many possible definitions that would be useful for different types of SV detection problems that vary in event type sought, depth of coverage of the experiment, heterogeneity of the sample, etc.

@@

We have implemented a deletion detection algorithm for medium-to-high coverage diploid samples in this framework called Cloudbreak. Our implementation defines the three application-specific functions described above as:
\begin{description}
\item[\sc{Loci}] Because we are detecting deletions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus. Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. We approximate the probability $p^{k}_c$ of each end alignment being correct by:

@@

We fit the parameters of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$. We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$. We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.

@@

We compared the performance of Cloudbreak for detecting deletions to a selection of popular tools: the RP method Breakdancer \autocite{Chen:2009p3}, GASVPro, an RP method that integrates RD signals and ambiguous mappings to increase sensitivity and specificity \autocite{Sindi:2012kk}, the SR method Pindel \autocite{Ye:2009p2}, and the hybrid RP-SR method DELLY \autocite{Rausch:2012he}. In each case we use default parameters and attempt to follow the recommend pipelines for alignment for each tool (See Methods for details).

We use the following criteria to define a true prediction given a gold standard set of deletions to test against: A predicted deletion is counted as a true positive if a) it overlaps with a deletion from the gold standard set, b) the length of the predicted call is within 300bp (the library fragment size in both our real and simulated libraries) of the length of the true deletion, and c) the true deletion has not been already been discovered by another prediction from the same method. Furthermore, we find that Pindel is very accurate at discovering deletions less than or equal to 40bp in length, so we exclude from consideration calls that match a true deletion of less than 40bp where the call is less than or equal to 75bp in length. We also discard all Pindel calls of less than or equal to 20bp. It should be noted that the methods tested here vary in their breakpoint resolution: SR methods have higher resolution than RP methods. Cloudbreak sacrifices additional resolution (by dividing the genome into 25bp windows); we strive however to increase sensitivity and specificity in the hopes that such calls could still be useful even if their resolution is limited. This seems reasonable to us given the possibility of pipelines in which RP calls are refined by SR mappings or local assembly methods.

@@

As has been observed elsewhere, there is no available test set of real Illumina sequencing data from a sample that has a complete annotation of structural variations from the reference. Therefore, testing with simulated data is important to fully characterize an algorithm's performance characteristics. On the other hand, it is important that the simulated data contain realistic SVs that follow patterns of SVs observed in real data. To address this, we took one of the most complete lists of SVs from a single sample available, the list of homozygous insertions and deletions from the genome of J. Craig Venter \autocite{Levy:2007fb}. Since there are relatively few heterozygous insertions and deletions annotated in the Venter genome, we randomly assigned each variant to be either homozygous or heterozygous and applied them to one or both of two copies of the human GRCh36 chromosome 2 reference sequence. We then simulated paired Illumina reads from these modified references using \emph{dwgsim} from the DNAA software package\footnote{\url{http://sourceforge.net/apps/mediawiki/dnaa/}}. We simulated 100bp reads with a mean fragment size of 300bp and a standard deviation of 30bp, and generated 15X coverage for each modified sequence. Pooling the reads from both simulations gives 30X coverage for a diploid sample with a mix of homozygous and heterozygous insertions and deletions.

\begin{figure}[t]
\centering
\includegraphics[width=0.75\textwidth]{CHR2SIM_DELS_ROC.pdf}
\caption{ROC curve of deletion prediction performance on a simulated set of reads giving diploid coverage of 30X on human chromosome 2 with insertions and deletions from the Venter genome added to one or both haplotypes. Thresholds vary by: Cloudbreak - likelihood ratio; Breakdancer, DELLY, GASVPro, and Pindel - number of supporting read pairs.}
\label{chr2roc}
\end{figure}
